Enforced Invariant Confluence

### 2018-05-01 doc/paper/sections/introduction.tex
minor tweakage, mostly for space

\section{Introduction}
When an application designer decides to replicate a piece of data, they have to
make a fundamental choice between weak and strong consistency. Replicating the
data with strong consistency using a technique like distributed transactions~\cite{bernstein1981concurrency,mohan1986transaction} or state machine
replication~\cite{schneider1990implementing, lamport1998part,
liskov2012viewstamped, ongaro2014search, moraru2013there, vincent2015designing,
oki1988viewstamped, lamport2005generalized, lamport2006fast} makes the
application designer's life very easy. To the application developer, a strongly
consistent system behaves exactly like a single-threaded system running on a
single node, so reasoning about the behavior of the system is
simple~\cite{herlihy1990linearizability}.  Unfortunately, strong consistency is
at odds with performance. The CAP theorem and PACELC theorem tell us that
strongly consistent systems suffer from higher latency at best and
unavailability at worst~\cite{gilbert2002brewer, brewer2012cap,
abadi2012consistency}. On the other hand, weak consistency models like eventual
consistency~\cite{vogels2009eventually}, PRAM
consistency~\cite{lipton1988pram}, causal consistency~\cite{ahamad1995causal},
and others~\cite{lloyd2011don, mehdi2017can} allow data to be replicated with
high availability and low latency, but they put a tremendous burden on the
application designer to reason about the complex interleavings of operations
that are allowed by these weak consistency models. In particular, weak consistency
models strip an application developer of one of the earliest and most effective
tools that is used to reason about the execution of programs: application
invariants~\cite{hoare1969axiomatic, balegas2015towards} and integrity constraints~\cite{IntegrityConstraintCitations}. Even if every
transaction executing in a weakly consistent system individually maintains an
application invariant, the system as a whole can produce invariant-violating
states.

Is it possible for us to have our strongly consistent cake and eat it with high
availability too? Can we replicate a piece of data with weak consistency but
still ensure that its invariants are maintained? Yes, sometimes. Bailis et al.\
introduced the notion of \emph{\invariantconfluence{}} as a necessary and sufficient
condition for when invariants can be maintained over replicated data without
the need for any coordination~\cite{bailis2014coordination}. If an object is
\invariantconfluent{} with respect to an invariant, we can replicate it with
the performance benefits of weak consistency and (some of) the correctness
benefits of strong consistency.

Unfortunately, to date, the task of identifying whether or not
an object actually is \invariantconfluent{} has remained an exercise in human proof generation. Bailis et al.\
manually categorized a set of common objects, transactions, and invariants
(e.g.\ foreign key constraints on relations, linear constraints on integers) as
\invariantconfluent{} or not. Hand-written proofs of this sort are unreasonable to expect from programmers. Ideally we would have a general purpose
program that can automatically determine \invariantconfluence{} for us.
\textbf{The first main thrust of this paper is to design a general-purpose
\invariantconfluence{} decision procedure, and implement it in an interactive system.}

Designing such a general-purpose decision procedure is
difficult\footnote{Actually, it's impossible. Determining whether an object is
\invariantconfluent{} is, in general, undecidable. Still, we can develop a
decision procedure that works well in the common case.}. Many existing
approaches instead develop a decision procedure for \emph{\invariantclosure{}}, a
sufficient (but not necessary) condition for
\invariantconfluence{}~\cite{li2012making, li2014automating}. The existing
approaches check to see if an object is \invariantclosed{}. If it is, then they
conclude that it is also \invariantconfluent{}. If it's not, then the current
approaches are unable to conclude anything about whether or not the object is
\invariantconfluent{}. In this paper, we take a step back and study the
underlying reason \emph{why} \invariantclosure{} is not necessary for
\invariantconfluence{}. We find that \invariantconfluence{} is fundamentally a
reachability property, checking whether certain values of our replicated
object can actually be obtained in the execution of a system. We discover that
unreachable, yet invariant satisfying, states cause \invariantclosure{} to be
an unnecessary condition. Using this understanding, we construct a set of
modest conditions under which \invariantclosure{} and \invariantconfluence{}
are in fact \emph{equivalent}, allowing us to reduce the problem of determining
\invariantconfluence{} to that of determining \invariantclosure{}.

@@

\textbf{The second main thrust of this paper is to take a step beyond
\invariantconfluence{} and develop a generalization of \invariantconfluence{}
called segmented \invariantconfluence{}.} While \invariantconfluence{}
characterizes objects that can be replicated \emph{without any} coordination,
segmented \invariantconfluence{} allows us to replicate
non-\invariantconfluent{} objects with only \emph{occasional coordination}. The main
idea is that invariant satisfying states are divided into a number of segments,
and the set of allowable transactions within each segment is restricted.
Within a segment, servers act without any coordination, synchronizing only to
transition across segment boundaries. This design highlights the trade-off
between application complexity and coordination-freedom; more complex
applications require more segments which require more coordination, and
vice-versa.

>>>>

\section{Introduction}
When an application designer decides to replicate a piece of data, they have to
make a fundamental choice between weak and strong consistency. Replicating the
data with strong consistency using a technique like distributed transactions (e.g., ~\cite{bernstein1981concurrency,mohan1986transaction}) or state machine
replication~\cite{schneider1990implementing, lamport1998part,
liskov2012viewstamped, ongaro2014search, moraru2013there, vincent2015designing,
oki1988viewstamped, lamport2005generalized, lamport2006fast} \jmh{why so many citations? pick a few and say "e.g."} makes the
application designer's life very easy. To the developer, a strongly
consistent system behaves exactly like a single-threaded system running on a
single node, so reasoning about the behavior of the system is
simple~\cite{herlihy1990linearizability}.  Unfortunately, strong consistency is
at odds with performance. The CAP theorem and PACELC theorem tell us that
strongly consistent systems suffer from higher latency at best and
unavailability at worst~\cite{gilbert2002brewer, brewer2012cap,
abadi2012consistency}. On the other hand, weak consistency models like eventual
consistency~\cite{vogels2009eventually}, PRAM
consistency~\cite{lipton1988pram}, causal consistency~\cite{ahamad1995causal},
and others~\cite{lloyd2011don, mehdi2017can} allow data to be replicated with
high availability and low latency, but they put a tremendous burden on the
application designer to reason about the complex interleavings of operations
that are allowed by these weak consistency models. In particular, weak consistency
models strip an application developer of one of the earliest and most effective
tools that is used to reason about the execution of programs: application
invariants~\cite{hoare1969axiomatic, balegas2015towards} such as database integrity constraints~\cite{IntegrityConstraintCitations}. Even if every
transaction executing in a weakly consistent system individually maintains an
application invariant, the system as a whole can produce invariant-violating
states.

Is it possible for us to have our strongly consistent cake and eat it with high
availability too? Can we replicate a piece of data with weak consistency but
still ensure that its invariants are maintained? Yes ... sometimes. Bailis et al.\
introduced the notion of \emph{\invariantconfluence{}} as a necessary and sufficient
condition for when invariants can be maintained over replicated data without
the need for any coordination~\cite{bailis2014coordination}. If an object is
\invariantconfluent{} with respect to an invariant, we can replicate it with
the performance benefits of weak consistency and (some of) the correctness
benefits of strong consistency.

Unfortunately, to date, the task of identifying whether or not
an object actually is \invariantconfluent{} has remained an exercise in human proof generation. Bailis et al.\
manually categorized a set of common objects, transactions, and invariants
(e.g.\ foreign key constraints on relations, linear constraints on integers) as
\invariantconfluent{} or not. Hand-written proofs of this sort are unreasonable to expect from programmers. Ideally we would have a general purpose
program that can automatically determine \invariantconfluence{} for us.
\textbf{The first main thrust of this paper is to make \invariantconfluence{} checkable:} to design a general-purpose
\invariantconfluence{} decision procedure, and implement it in an interactive system.

Designing such a general-purpose decision procedure is
difficult\footnote{In fact it's impossible. Determining 
\invariantconfluence{} of an object is undecidable in general. Still, we can develop a
decision procedure that works well in the common case.}. Many prior
efforts instead develop a decision procedure for \emph{\invariantclosure{}}, a
sufficient (but not necessary) condition for
\invariantconfluence{}~\cite{li2012making, li2014automating}. The existing
approaches check whether an object is \invariantclosed{}. If it is, then they
conclude that it is also \invariantconfluent{}. If it's not, then the current
approaches are unable to conclude anything about whether or not the object is
\invariantconfluent{}. In this paper, we take a step back and study the
underlying reason \emph{why} \invariantclosure{} is not necessary for
\invariantconfluence{}. We find that \invariantconfluence{} is fundamentally a
reachability property, checking whether certain values of our replicated
object can actually be obtained in the execution of a system. We discover that
unreachable, yet invariant satisfying, states cause \invariantclosure{} to be
an unnecessary condition. Using this understanding, we construct a set of
modest conditions under which \invariantclosure{} and \invariantconfluence{}
are in fact \emph{equivalent}, allowing us to reduce the problem of determining
\invariantconfluence{} to that of determining \invariantclosure{}.

@@

\textbf{The second main thrust of this paper is to partially avoid coordination even in programs that require it}, by generalizing \invariantconfluence{} to a local property called \emph{segmented \invariantconfluence{}} over subsets of program states. While \invariantconfluence{}
characterizes objects that can be replicated \emph{without any} coordination,
segmented \invariantconfluence{} allows us to replicate
non-\invariantconfluent{} objects with only \emph{occasional} coordination. The main
idea is to divide the set of invariant-satisfying states into \emph{segments}, with a restricted set of transactions allowed in each segment.
Within a segment servers act without any coordination; they synchronize only to
transition across segment boundaries. This design highlights the trade-off
between application complexity and coordination-freedom; more complex
applications require more segments which require more coordination, and
vice-versa.
